Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-11-12 manuscript/manuscript.tex
link sections to notebooks 01-08

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.

Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (Notebook 00).
Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that properties that we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (Notebook 00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

@@

TICA yields a representation of our molecular simulation data with a reduced dimensionality,
which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
which is shown in the showcase notebook (00).

@@

@@

@@

@@

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.

>>>>

In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
In order to approximate the slow dynamics in a statistically efficient manner,
a lower dimensional representation of our simulation data is necessary.
However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
provide a systematic means to quantitatively compare multiple representations of the simulation data.
In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
In Notebook (01), we present in detail how to extract features from MD datasets and how to systematically compare them.

Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features reflecting protein structure,
we compute the (cross-validated) VAMP-2 score (Notebook 00).
Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
it is important to ensure that properties that we optimize are robust as a function of lag time. 
Consequently, we compute the VAMP-2 score at several lag times (Notebook 00). 
We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b. 
We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
This suggests that backbone torsions are the best of the options evaluated for MSM construction.

@@

We demonstrate how to apply TICA, suggest how to interpret the projected coordinates and compare the results to other dimension reduction techniques in Notebook (02).

@@

TICA yields a representation of our molecular simulation data with a reduced dimensionality,
which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
which is shown in the showcase Notebook (00).
A detailed comparison between different clustering techniques is provided in Notebook (02).

@@

In Notebook (03), we demonstrate in detail how to estimate and validate MSMs with PyEMMA.

@@

More details about (spectral) properties of MSMs and how to analyze them with PyEMMA are discussed in Notebook (04) and Notebook (05).

@@

Besides a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.

@@

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
We further demonstrate the symptoms of difficult data situations and how to deal with them in Notebook (08).
